Paper

### 2025-03-01 paper/paper.qmd
paper update

An agent-based model (ABM) was implemented using NetLogo [@wilenskyu.1999]. Briefly, the simulation started by building a virtual hospital with a configurable high-level structure, defined by its number of wards, beds per ward, and proportion of bedspaces sharing toilet facilities. Next, the hospital was populated with patients, each having an infected/colonised or non-infected/colonised state, defined as:

@@

Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgado2009]. Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].

## Model calibration

@@

The approach to calibration was sequential. Firstly, ABC was used to calibrate the model pre-outbreak data to identify parameters that simulated the baseline rate. Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgado2009]. To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset. The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency. At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures. The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.

The posterior parameter distributions were used to simulate 20 outbreak scenarios. Parameter samples were drawn from a uniform distribution:

@@

: Parameters that were assumed to be constant during calibration. {#tbl-priors}

@@

## Outbreak generation

@@

@@

```{r, echo=FALSE}
posteriors <- abc_params_outbreak_control$unadj.values
comm_colonisation <- round(mean(posteriors[, "community-colonisation-rate"] * 100), 2)
outbreak_colonisation <- round(mean(posteriors[, "o-community-colonisation-rate"] * 100), 2)
outbreak_antibiotic_rate <- round(mean(posteriors[, "o-antibiotic-prescription-rate"] * 100), 2)
control_antibiotic_rate <- round(mean(posteriors[, "c-antibiotic-prescription-rate"] * 100), 2)
```

```{r, include=T}
#| tbl-cap: "Summary of parameter posteriors after calibration using Approximate bayesian computation."
#| label: tbl-posteriors
#| echo: false
params_table <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble() %>% 
  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
  summarise(across(everything(), ~ paste0(
    round(mean(.x), 2),
    " (",
    round(sd(.x), 2),
    ")"
    )))

param_units <- c("toilet-frequenting-rate" = "times per day",
                 "toilet-cleaning-rate" = "times per day",
                 "toilet-contamination-effect" = "",
                 "toilet-cleaning-effect" = "",
                 "community-colonisation-rate" = "proportion of patients",
                 "antibiotic-prescription-rate" = "proportion of admissions",
                 "antibiotic-effect" = "relative risk",
                 "random-colonisation" = "cases per 10,000 bed days")

param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
                 "toilet-cleaning-rate" = "Toilet cleaning rate",
                 "toilet-contamination-effect" = "Toilet contamination effect",
                 "toilet-cleaning-effect" = "Toilet cleaning effect",
                 "community-colonisation-rate" = "Community colonisation rate",
                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                 "antibiotic-effect" = "Antibiotic effect",
                 "random-colonisation" = "Random colonisation")

@@

After calibration, key parameters included: community colonisation rate of `r comm_colonisation`%, rising to `r outbreak_colonisation`% during the outbreak period. Furthermore, the proportion of patients prescribed an antimicrobial during their admission increased from `r consts[["antibiotic-prescription-rate"]] * 100`% to `r outbreak_antibiotic_rate`% during the outbreak; then reduced to `r control_antibiotic_rate`% during the period of enhanced infection control measures. The posterior parameter distributions are shown in @fig-posterior-densities. Simulated data had a pre-outbreak baseline median rate of `r round(pre_outbreak_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\]. During the outbreak period, simulations had a median rate of `r round(outbreak_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).

```{r}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| warning: false
#| label: fig-posterior-densities
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
#| fig-subcap:
#|   - "General parameters"
#|   - "Toilet-related parameters"
#| layout-ncol: 1
abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))

@@

```{r}
#| echo: false
#| label: fig-simulated-rates
#| fig.cap: "Observed and simulated rates of _C. difficile_ infection over the simulated period. Approximate Bayesian computation was used to determine the unobserved parameters of the agent-based model. The red line shows the true observed infection rates. The black solid line shows the mean simulated rates; the dashed lines show the standard deviation."
#| fig.width: 18
#| fig.height: 8
ggplot(long_observed_ss, aes(x = x, y = rates)) + geom_line(color = 'red') +
  geom_line(data = sim_mean, aes(x = x, y = m)) +
  geom_line(data = sim_mean, aes(x = x, y = q_high), linetype='dashed') +
  geom_line(data = sim_mean, aes(x = x, y = q_low), linetype='dashed') + 
  theme(text=element_text(size=24)) +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

```{r}
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| include: false
#| message: false
consts_sim_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  #slice_sample(n=10) %>% 
  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(siminputrow, date_sim) %>%
  summarise(rate = mean(rate)) %>%
  ggplot(aes(x = date_sim,
             y = rate,
             color = factor(`siminputrow`))) +
  geom_line() +
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
  guides(color="none")
```

```{r}
single_outbreak_sim_mod <- single_outbreak_sim %>% 
  mutate(prop_infected = `current-colonised` / `current-inpatients`,
         current_not_infected = `current-inpatients` - `current-colonised`)
single_outbreak_sim_mod %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  ggplot(aes(x = `[step]`, y = value, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment")

@@

```{r, cache=TRUE}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

@@

output_df <- filter(output_df, time > 31)

# Plot the results
library(ggplot2)
output_long <- reshape2::melt(output_df, id = "time")

>>>>

An agent-based model (ABM) was implemented using NetLogo [@wilenskyu.NetLogo1999]. Briefly, the simulation started by building a virtual hospital with a configurable high-level structure, defined by its number of wards, beds per ward, and proportion of bedspaces sharing toilet facilities. Next, the hospital was populated with patients, each having an infected/colonised or non-infected/colonised state, defined as:

@@

Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgadoAnalysisOutbreakClostridium2009]. Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].

## Model calibration {#sec-model-calibration}

@@

The approach to calibration was sequential. Firstly, ABC was used to calibrate the model pre-outbreak data to identify parameters that simulated the baseline rate. Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009]. To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset. The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency. At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures. The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.

The posterior parameter distributions were used to simulate `r length(unique(consts_sim$siminputrow))` outbreak scenarios. Parameter samples were drawn from a uniform distribution:

@@

: Parameter prior distributions or constants. {#tbl-priors}

@@

## Outbreak generation {#sec-outbreak-generation}

@@

## Sensitivity analysis {#sec-sens-analysis}

```{r, include=FALSE}
sens_n_seeds <- length(unique(consts_sim$siminputrow))
```

To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019]. Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) proportion of patients redistributed to a single-occupancy bed after colonisation/infection; 3) toilet cleaning rate; and 4) toilet cleaning effect. To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection ABC. Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value. The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period. For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.

@@

```{r, include=FALSE}
posteriors <- abc_params_outbreak_control$unadj.values
comm_colonisation <- round(mean(posteriors[, "community-colonisation-rate"] * 100), 2)
outbreak_colonisation <- round(mean(posteriors[, "o-community-colonisation-rate"] * 100), 2)
outbreak_antibiotic_rate <- round(mean(posteriors[, "o-antibiotic-prescription-rate"] * 100), 2)
control_antibiotic_rate <- round(mean(posteriors[, "c-antibiotic-prescription-rate"] * 100), 2)
b_toilet_clean_rate <- round(mean(posteriors[, "toilet-cleaning-rate"]), 2)
o_toilet_clean_rate <- round(mean(posteriors[, "o-toilet-cleaning-rate"]), 2)
c_toilet_clean_rate <- round(mean(posteriors[, "c-toilet-cleaning-rate"]), 2)
```

An agent-based model that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced. After calibrating the agent-based model using approximate Bayesian computation (ABC), three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control. A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection ABC) is shown in @tbl-posteriors. The densities of the posterior parameter distributions are shown in @fig-posterior-densities. In line with the prior bounds described in @sec-outbreak-generation, the posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).

```{r tbl-posteriors, include=TRUE}
#| tbl-cap: "Summary of parameter posteriors after calibration using Approximate bayesian computation."
#| echo: false
params_table <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble() %>% 
  mutate("community-colonisation-rate" = `community-colonisation-rate` * 100,
         "o-community-colonisation-rate" = `o-community-colonisation-rate` * 100,
         "o-antibiotic-prescription-rate" = `o-antibiotic-prescription-rate` * 100,
         "c-antibiotic-prescription-rate" = `c-antibiotic-prescription-rate` * 100) %>% 
  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
  summarise(across(everything(), ~ paste0(
    round(mean(.x), 2),
    " (",
    round(sd(.x), 2),
    ")"
    )))

param_units <- c("toilet-frequenting-rate" = "times per day",
                 "toilet-cleaning-rate" = "times per day",
                 "toilet-contamination-effect" = "",
                 "toilet-cleaning-effect" = "",
                 "community-colonisation-rate" = "percentage of patients",
                 "antibiotic-prescription-rate" = "percentage of admissions",
                 "antibiotic-effect" = "relative risk",
                 "random-colonisation" = "cases per 10,000 bed days",
                 "proportion-redistributed" = "proportion of patients")

param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
                 "toilet-cleaning-rate" = "Toilet cleaning rate",
                 "toilet-contamination-effect" = "Toilet contamination effect",
                 "toilet-cleaning-effect" = "Toilet cleaning effect",
                 "community-colonisation-rate" = "Community colonisation rate",
                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                 "antibiotic-effect" = "Antibiotic effect",
                 "random-colonisation" = "Random colonisation",
                 "proportion-redistributed" = "Proportion redistributed")

@@

```{r fig-posterior-densities, include=TRUE}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| warning: false
#| label: fig-posterior-densities
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
#| fig-subcap:
#|   - "General parameters"
#|   - "Toilet-related parameters"
#| layout-ncol: 1
abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))

@@

Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-simulated-rates). In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\]. During the outbreak period, simulations had a median rate of `r round(outbreak_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).

```{r fig-simulated-rates, include=TRUE}
#| echo: false
#| label: fig-simulated-rates
#| fig.cap: "Observed and simulated rates of _C. difficile_ infection over the simulated period. Approximate Bayesian computation was used to determine the unobserved parameters of the agent-based model. The red line shows the true observed infection rates. The black solid line shows the mean simulated rates; the dashed lines show the standard deviation."
#| fig.width: 18
#| fig.height: 8
ggplot(long_observed_ss, aes(x = x, y = rates)) + geom_line(color = 'red') +
  geom_line(data = sim_mean, aes(x = x, y = m)) +
  geom_line(data = sim_mean, aes(x = x, y = q_high), linetype='dashed') +
  geom_line(data = sim_mean, aes(x = x, y = q_low), linetype='dashed') + 
  theme(text=element_text(size=24)) +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

```{r, include=FALSE}
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| message: false
consts_sim_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  #slice_sample(n=10) %>% 
  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(siminputrow, date_sim) %>%
  summarise(rate = mean(rate)) %>%
  ggplot(aes(x = date_sim,
             y = rate,
             color = factor(`siminputrow`))) +
  geom_line() +
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
  guides(color="none")
```

```{r, include=FALSE}
single_outbreak_sim_mod <- single_outbreak_sim %>% 
  mutate(prop_infected = `current-colonised` / `current-inpatients`,
         current_not_infected = `current-inpatients` - `current-colonised`)
single_outbreak_sim_mod %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  ggplot(aes(x = `[step]`, y = value, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment")

@@

The sensitivity analysis explored the impact of different infection control practices on the number of cases observed during the outbreak period -- the results are shown in @fig-sensitivity-analysis. The analysis showed that the cleaning of toilets had a dramatic impact on the number of cases observed during the outbreak period, although the benefit appeared to plateaeu beyond a certain amount -- for example, cleaning toilets more than twice per day appeared to have marginal benefits. The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate.

```{r fig-sensitivity-analysis, include=TRUE}
#| echo: false
list(abx_prescription_rate = sens_abx_prescription_outbreak_cases,
     proportion_redistributed = sens_prop_redist_outbreak_cases,
     toilet_cleaning_rate = sens_toilet_cleaning_rate_outbreak_cases,
     toilet_cleaning_effect = sens_toilet_cleaning_effect_outbreak_cases) %>%
  bind_rows() %>% 
  pivot_longer(cols = starts_with("c-"),
               names_to = "param",
               values_drop_na = TRUE) %>% 
  group_by(param, value) %>% 
  summarise(mean_cases = mean(`outbreak-cases`),
            sd_cases = sd(`outbreak-cases`),
            .groups = "keep") %>% 
  ggplot(aes(x = value, y = mean_cases)) + 
  geom_line() + 
  # shade sd
  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
  facet_wrap(~ param, scales = "free")
```

```{r}
sens_prop_redist_perfect_clean_outbreak_cases %>% pivot_longer(cols = `c-proportion-redistributed`,
               names_to = "param",
               values_drop_na = TRUE) %>% 
  group_by(param, value) %>% 
  summarise(mean_cases = mean(`outbreak-cases`),
            sd_cases = sd(`outbreak-cases`),
            .groups = "keep") %>% 
  ggplot(aes(x = value, y = mean_cases)) + 
  geom_line() + 
  # shade sd
  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
  facet_wrap(~ param, scales = "free")
```

```{r run-deterministic-model, include=FALSE, cache=TRUE}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

@@

output_df <- filter(output_df, time > 31)
```

```{r fig-single-outbreak, include=TRUE}
#| echo: false
output_long <- reshape2::melt(output_df, id = "time")
